---
title: "ALS_initial_research"
author: "Alex Svetov & Tetyana Yatsenko"
date: "2023-09-10"
output: html_document
---

```{r setup, include=FALSE}
if(!require(knitr)) install.packages("knitr")

knitr::opts_chunk$set(echo = TRUE)

if(!require(dplyr)) install.packages("dplyr")
if(!require(ggplot2)) install.packages("ggplot2")
if(!require(xtable)) install.packages("xtable")
if(!require(devtools)) install.packages("devtools")
if(!require(rstudioapi)) install.packages("rstudioapi")

library(dplyr)
library(devtools)
library(ggplot2)
library(xtable)
library(rstudioapi) 
```

## Read ALSFRS Data, ALSFRS hisrory, death data, riluzote data
```{r}
workdir <- dirname(rstudioapi::getActiveDocumentContext()$path)
setwd(workdir) 

finenameALSFRS <- "PROACT_ALSFRS.csv"
finenameALSFRSHistory <- "PROACT_ALSHISTORY.csv"
finenameDeath <- "PROACT_DEATHDATA.csv"
finenameRiluzole <- "PROACT_RILUZOLE.csv"

alsrfs <-  read.csv(file = paste(workdir,finenameALSFRS,sep="/"))
alsrfs$subject_id <- as.factor(alsrfs$subject_id)
print(str(alsrfs))

alsrfsH <-  read.csv(file = paste(workdir,finenameALSFRSHistory,sep="/"))
alsrfsH$subject_id <- as.factor(alsrfsH$subject_id)
print(str(alsrfsH))

deathdata <-  read.csv(file = paste(workdir,finenameDeath,sep="/"))
deathdata$subject_id <- as.factor(deathdata$subject_id)
print(str(deathdata))

riluzole <-  read.csv(file = paste(workdir,finenameRiluzole,sep="/"))
riluzole$subject_id <- as.factor(riluzole$subject_id)
print(str(riluzole))

columnsALSFRS1 <- c("subject_id","ALSFRS_Delta","ALSFRS_R_Total")
columnsALSFRS2 <- c("subject_id","ALSFRS_Delta","avg_ALSFRS_R_Total")
columnsALSFRSH <- c("subject_id",	"Subject_ALS_History_Delta", "Onset_Delta", "Diagnosis_Delta")

```

## Clean death, riluzole data
```{r}
## Death Data
# subject_id is not unique here. 
# it is one person with two death date, i will choose the last
f_deathdata <- 
  deathdata  %>% 
  unique() %>%
  group_by(subject_id,Subject_Died) %>%
  summarise(Death_Days_max = max(Death_Days)) %>%
  ungroup()

colnames(f_deathdata)[colnames(f_deathdata) == "Death_Days_max"] = "Death_Days"
print(str(f_deathdata))

## Riluzole  Data
# subject_id is unique here. I have checked
f_riluzole <- 
  riluzole  %>% 
  unique() 

print(str(f_riluzole))

```



## Clean, checking and creation features ALSFRS hisrory
```{r}
## ALSRFRS Historical  Data
# ALSRFRS Historical data
f_alsrfsH <- 
  alsrfsH %>% 
  select(all_of(columnsALSFRSH)) 

# Unique subject_id alsrfsHistory 
print(length(unique(f_alsrfsH$subject_id)))

# Data set with Onset_delta
f_alsrfsH <- 
  f_alsrfsH %>% 
  filter(!is.na(Onset_Delta)) %>%
  unique() 

f_alsrfsH$Onset_weeks  <- round( f_alsrfsH$Onset_Delta/ 7.0 )
f_alsrfsH$Onset_months <-round( f_alsrfsH$Onset_Delta/ 30.0 )

print(str(f_alsrfsH))

# Data set with Onset_delta & Diagnosis_Delta
f_alsrfsH <- 
  f_alsrfsH %>% 
  filter(!is.na(Diagnosis_Delta)) %>%
  unique() 

f_alsrfsH$Diagnosis_weeks <- round( f_alsrfsH$Diagnosis_Delta/ 7.0 )
f_alsrfsH$Diagnosis_months <- round(f_alsrfsH$Diagnosis_Delta/ 30.0 )

# time difference between diagnosis time and onset time
f_alsrfsH$OnsetDiag_days <- 
  f_alsrfsH$Diagnosis_Delta - f_alsrfsH$Onset_Delta

f_alsrfsH$OnsetDiag_weeks <- round( f_alsrfsH$OnsetDiag_days/ 7.0 )
f_alsrfsH$OnsetDiag_months <- round( f_alsrfsH$OnsetDiag_days/ 30.0 )

print(str(f_alsrfsH))


```


##  Add Death, riluzole Data ALSRFS History 
```{r}

#ToDo inner or left join with Death Data

f_alsrfsH <-
  f_alsrfsH  %>%
  inner_join(f_deathdata, by=c("subject_id" = "subject_id")) %>%
  left_join(f_riluzole, by=c("subject_id" = "subject_id"))

f_alsrfsH$Death_weeks <- round( f_alsrfsH$Death_Days/ 7.0 )
f_alsrfsH$Death_months <- round(f_alsrfsH$Death_Days/ 30.0 )

f_alsrfsH$Riluzole_use_weeks <- round( f_alsrfsH$Riluzole_use_Delta/ 7.0 )
f_alsrfsH$Riluzole_use_months <- round(f_alsrfsH$Riluzole_use_Delta/ 30.0 )

# difference between death time and onset time
f_alsrfsH$DeathOnset_days <- f_alsrfsH$Death_Days - f_alsrfsH$Onset_Delta
f_alsrfsH$DeathOnset_weeks <- round( f_alsrfsH$DeathOnset_days/ 7.0 )
f_alsrfsH$DeathOnset_months <- round(f_alsrfsH$DeathOnset_days/ 30.0 )

# difference between death time and diagnosis time
f_alsrfsH$DeathDiag_days <- f_alsrfsH$Death_Days - f_alsrfsH$Diagnosis_Delta
f_alsrfsH$DeathDiag_weeks <- round( f_alsrfsH$DeathDiag_days/ 7.0 )
f_alsrfsH$DeathDiag_months <- round(f_alsrfsH$DeathDiag_days/ 30.0 )

# difference between death time and Riluzole use time
f_alsrfsH$DeathRiluzole_days <- f_alsrfsH$Death_Days - f_alsrfsH$Riluzole_use_Delta
f_alsrfsH$DeathRiluzole_weeks <- round( f_alsrfsH$DeathRiluzole_days/ 7.0 )
f_alsrfsH$DeathRiluzole_months <- round(f_alsrfsH$DeathRiluzole_days/ 30.0 )

# difference between Diagnosis time and Riluzole use time
f_alsrfsH$DiagRiluzole_days <- f_alsrfsH$Diagnosis_Delta - f_alsrfsH$Riluzole_use_Delta
f_alsrfsH$DiagRiluzole_weeks <- round( f_alsrfsH$DiagRiluzole_days/ 7.0 )
f_alsrfsH$DiagRiluzole_months <- round(f_alsrfsH$DiagRiluzole_days/ 30.0 )

print(str(f_alsrfsH))
```


##  Cleanig of ALSRFRS  Data
```{r}
##  -------------
## ALSRFRS  Data
# select columns "subject_id","ALSFRS_Delta","ALSFRS_R_Total"
# filter rows with values in columns "ALSFRS_Delta","ALSFRS_R_Total"
# average value of "ALSFRS_R_Total" in a certain in moment of time
# we have several records some time


f_alsrfs <- 
  alsrfs %>% 
  select(all_of(columnsALSFRS1)) %>% 
  filter(!is.na(ALSFRS_R_Total)) %>%
  filter(!is.na(ALSFRS_Delta))  %>%
  group_by(subject_id, ALSFRS_Delta) %>%
  summarise(avg_ALSFRS_R_Total = mean(ALSFRS_R_Total)) %>%
  select(all_of(columnsALSFRS2)) %>%
  ungroup() 


colnames(f_alsrfs)[colnames(f_alsrfs) == "avg_ALSFRS_R_Total"] = "ALSFRS_R_Total"
f_alsrfs$ALSFRS_R_Total <- round( f_alsrfs$ALSFRS_R_Total )

print(str(f_alsrfs))
```
##  Original statistics preparation for ALSFRS  Data
```{r}
## 
# Calculate simple statistics: number of measurement (visits), day of first visit, last visit
# time between first and last visit days, weeks, months
sf_alsrfs <- 
  f_alsrfs %>%
  group_by(subject_id) %>%
  summarise( Number_visits = n(), 
             first_ALSFRS = min(ALSFRS_Delta), 
             last_ALSFRS = max(ALSFRS_Delta), 
             interval_ALSFRS = 1+max(ALSFRS_Delta) - min(ALSFRS_Delta) ) 

# time of first & last ALSFRS
sf_alsrfs$first_ALSFRS_weeks <- round( sf_alsrfs$first_ALSFRS / 7.0 )
sf_alsrfs$first_ALSFRS_months <- round( sf_alsrfs$first_ALSFRS / 30.0 )
sf_alsrfs$last_ALSFRS_weeks <- round( sf_alsrfs$last_ALSFRS / 7.0 )
sf_alsrfs$last_ALSFRS_months <- round( sf_alsrfs$last_ALSFRS / 30.0 )
sf_alsrfs$interval_ALSFRS_weeks <- round( sf_alsrfs$interval_ALSFRS / 7.0 )
sf_alsrfs$interval_ALSFRS_months <- round( sf_alsrfs$interval_ALSFRS / 30.0 )
# Add value of ALSFRS_R in the first measurement
sf_alsrfs <- 
  sf_alsrfs  %>%
  inner_join(f_alsrfs, by=c("subject_id" = "subject_id", "first_ALSFRS" =  "ALSFRS_Delta")) %>%
  unique()

# the value of the first ALSFRS
colnames(sf_alsrfs)[colnames(sf_alsrfs) == "ALSFRS_R_Total"] = "intial_ALSFRS_R_Total"

# Add value of ALSFRS_R in the last measurement
sf_alsrfs <- 
  sf_alsrfs  %>%
  inner_join(f_alsrfs, by=c("subject_id" = "subject_id", "last_ALSFRS" =  "ALSFRS_Delta")) %>%
  unique()

# the value of the last ALSFRS
colnames(sf_alsrfs)[colnames(sf_alsrfs) == "ALSFRS_R_Total"] = "last_ALSFRS_R_Total"

print(str(sf_alsrfs))
```
##  Intersection ALSRFS History with Onset&Diag Data with Statistic ALSRFS
```{r}
## Inner intersection
## ALSRFS History, Dead info, with table ALSRFS value
# Necessary condition: more that one ALSFRS_R

sf_alsrfsH <-
  sf_alsrfs  %>%
  inner_join(f_alsrfsH, by=c("subject_id" = "subject_id")) %>%
  filter(Number_visits>1)  

print(dim(sf_alsrfs))
print(dim(f_alsrfsH))
print(dim(sf_alsrfsH))

# time difference between time of first ALSFRS and Onset time
sf_alsrfsH$OnsetFirstALSFRS_days <- sf_alsrfsH$first_ALSFRS - sf_alsrfsH$Onset_Delta
sf_alsrfsH$OnsetFirstALSFRS_weeks <- round(sf_alsrfsH$OnsetFirstALSFRS_days /7.0)
sf_alsrfsH$OnsetFirstALSFRS_months <- round(sf_alsrfsH$OnsetFirstALSFRS_days/30.0)

# time difference between time of first ALSFRS and diagnosis time 
sf_alsrfsH$DiagFirstALSFRS_days <- sf_alsrfsH$first_ALSFRS - sf_alsrfsH$Diagnosis_Delta
sf_alsrfsH$DiagFirstALSFRS_weeks <- round(sf_alsrfsH$DiagFirstALSFRS_days /7.0)
sf_alsrfsH$DiagFirstALSFRS_months <- round(sf_alsrfsH$DiagFirstALSFRS_days/30.0)

# time difference between time of death and last ALSFRS time 
sf_alsrfsH$DeathLastALSFRS_days <- sf_alsrfsH$Death_Days - sf_alsrfsH$last_ALSFRS
sf_alsrfsH$DeathLastALSFRS_weeks <- round(sf_alsrfsH$DeathLastALSFRS_days /7.0)
sf_alsrfsH$DeathLastALSFRS_months <- round(sf_alsrfsH$DeathLastALSFRS_days/30.0)

# time difference between time of death and Diagnostics
sf_alsrfsH$DeathDiag_days <- sf_alsrfsH$Death_Days - sf_alsrfsH$Diagnosis_Delta
sf_alsrfsH$DeathDiag_weeks <- round(sf_alsrfsH$DeathDiag_days /7.0)
sf_alsrfsH$DeathDiag_months <- round(sf_alsrfsH$DeathDiag_days/30.0)

# time difference between time of death and Onset
sf_alsrfsH$DeathOnset_days <- sf_alsrfsH$Death_Days - sf_alsrfsH$Onset_Delta
sf_alsrfsH$DeathOnset_weeks <- round(sf_alsrfsH$DeathOnset_days /7.0)
sf_alsrfsH$DeathOnset_months <- round(sf_alsrfsH$DeathOnset_days/30.0)

# time difference between time of last ALSFRS time and Diagnostics
sf_alsrfsH$LastALSFRSDiag_days <- sf_alsrfsH$last_ALSFRS - sf_alsrfsH$Diagnosis_Delta
sf_alsrfsH$LastALSFRSDiag_weeks <- round(sf_alsrfsH$LastALSFRSDiag_days /7.0)
sf_alsrfsH$LastALSFRSDiag_months <- round(sf_alsrfsH$LastALSFRSDiag_days/30.0)

# time difference between time of last ALSFRS time and Onset
sf_alsrfsH$LastALSFRSOnset_days <- sf_alsrfsH$last_ALSFRS - sf_alsrfsH$Onset_Delta
sf_alsrfsH$LastALSFRSOnset_weeks <- round(sf_alsrfsH$LastALSFRSOnset_days /7.0)
sf_alsrfsH$LastALSFRSOnset_months <- round(sf_alsrfsH$LastALSFRSOnset_days/30.0)

print(str(sf_alsrfsH))
```
##  Intial ALSFRS_R
```{r}
  #Statistics for intial_ALSFRS_R
  print(dim(sf_alsrfsH))
  print(summary(sf_alsrfsH$intial_ALSFRS_R_Total))
  print(tapply(sf_alsrfsH$intial_ALSFRS_R_Total,sf_alsrfsH$Subject_Died, summary))
 
  #Categories of initial ALSRFS-R 
   sf_alsrfsH <-
    sf_alsrfsH %>%
    mutate(type=case_when(
      .$intial_ALSFRS_R_Total >=41  ~ "41>=",
      .$intial_ALSFRS_R_Total <41 & .$intial_ALSFRS_R_Total >= 38  ~ "38>=,<41",
      .$intial_ALSFRS_R_Total < 38 & .$intial_ALSFRS_R_Total >= 34  ~ "34>=,<38",
      .$intial_ALSFRS_R_Total < 34 ~ "34<")) 
   
  # Statistics for types
  print(dim(sf_alsrfsH))
  print(summary(sf_alsrfsH$intial_ALSFRS_R_Total))
  print(tapply(sf_alsrfsH$intial_ALSFRS_R_Total,sf_alsrfsH$type, summary))
   
```

## Slope of Degradation
```{r}
# Calculate slope of degradation

  sf_alsrfsH<- 
    sf_alsrfsH %>%
    filter(interval_ALSFRS_months>0) %>%
    mutate(diff_ALSFRS=last_ALSFRS_R_Total-intial_ALSFRS_R_Total) %>%
    mutate(slope_ALSFRS=diff_ALSFRS / interval_ALSFRS_months) 


# Statistics for slope
  print(dim(sf_alsrfsH))
  print(summary(sf_alsrfsH$slope_ALSFRS))
  print(tapply(sf_alsrfsH$slope_ALSFRS,sf_alsrfsH$type, summary))

print(dim(sf_alsrfsH))
```
##  Diagnosis Time
```{r}
 #Statistics for Diagnosis_months 
 print(dim(sf_alsrfsH))
 print(summary(sf_alsrfsH$Diagnosis_months))
 print(tapply(sf_alsrfsH$Diagnosis_months,sf_alsrfsH$Subject_Died, summary))
 
#Categories of Diagnosis months
 sf_alsrfsH <-
    sf_alsrfsH %>%
    mutate(sf_alsrfsH,typeDiagTime=case_when(
      .$Diagnosis_months >=-2.0  ~ ">=-2.0",
      .$Diagnosis_months < -2.0 &  .$Diagnosis_months >= -5.0 ~ ">=-5.0,<-2.0",
      .$Diagnosis_months < -5.0 &  .$Diagnosis_months >= -10.0 ~ ">=-10.0,<-5.0",
      .$Diagnosis_months < -10.0  ~ "<-10")) 
 
```
##  Categories information
```{r}
  sf_alsrfsH <- 
    sf_alsrfsH %>% 
    mutate(DiedYes=case_when(
      .$Subject_Died =="Yes"  ~ 1,
      .$Subject_Died =="No"  ~ 0)) %>% 
    mutate(DiedNo=case_when(
      .$Subject_Died =="Yes"  ~ 0,
      .$Subject_Died =="No"  ~ 1))

  type_sf_alsrfsH <-
    sf_alsrfsH %>%
    group_by(type) %>% 
    summarise(typeSize = n(),
              typeSizeDiedYes=sum(DiedYes),
              typeSizeDiedNo=sum(DiedNo)
              ) %>% 
         ungroup()
    
  print(type_sf_alsrfsH)
```

##  Add patient information information to traectories

```{r}

f_alsrfs$ALSFRS_Delta_weeks <- round(f_alsrfs$ALSFRS_Delta / 7.0 )
f_alsrfs$ALSFRS_Delta_months <- round(f_alsrfs$ALSFRS_Delta / 30.0 )
col <- c("subject_id","type","Number_visits","Subject_Died","typeDiagTime")

f_alsrfs <-
    f_alsrfs  %>%
    inner_join(select(sf_alsrfsH,all_of(col)), by=c("subject_id" = "subject_id"))
  
print(str(f_alsrfs))
  
```
## Traectories of degradations
```{r pressure, echo=FALSE}
  tr<- ggplot(f_alsrfs, aes(x = ALSFRS_Delta_months, y = ALSFRS_R_Total,group = subject_id)) +
    geom_line(aes(color =Subject_Died),size=1)+
    ylab("ALSFRS_R") +
    xlab("Time of measurement of ALSFRS_R, Months")+
    labs(title="Traectories of Degradations")
  tr + facet_grid(typeDiagTime ~type , margins = TRUE, scales='free', labeller=label_both)
  

```