---
title: "ALS_initial_research"
author: "Tetyana Yatsenko"
date: "2023-03-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if(!require(dplyr)) install.packages("dplyr")
if(!require(ggplot2)) install.packages("ggplot2")
if(!require(devtools)) install.packages("devtools")

library(dplyr)
library(devtools)
library(ggplot2)
```

## Read ALS Data


```{r}

workdir <- getwd() 

finenameALSFRS <- "PROACT_ALSFRS.csv"
finenameDeath <- "PROACT_DEATHDATA.csv"
finenameRiluzole <- "PROACT_RILUZOLE.csv"


alsrfs <-  read.csv(file = paste(workdir,finenameALSFRS,sep="/"))
alsrfs$subject_id <- as.factor(alsrfs$subject_id)
print(str(alsrfs))

deathdata <-  read.csv(file = paste(workdir,finenameDeath,sep="/"))
deathdata$subject_id <- as.factor(deathdata$subject_id)
print(str(deathdata))

riluzole <-  read.csv(file = paste(workdir,finenameRiluzole,sep="/"))
riluzole$subject_id <- as.factor(riluzole$subject_id)
print(str(riluzole))



impotantcolumns1 <- c("subject_id","ALSFRS_Delta","ALSFRS_R_Total")
impotantcolumns2 <- c("subject_id","ALSFRS_Delta","avg_ALSFRS_R_Total")
```

## Data mining
```{r}

# it is one person with two death date, i will choose the last
filter_deathdata <- 
  deathdata  %>% 
  unique() %>%
  group_by(subject_id,Subject_Died) %>%
  summarise(Death_Days_max = max(Death_Days)) %>%
  ungroup()

colnames(filter_deathdata)[colnames(filter_deathdata) == "Death_Days_max"] = "Death_Days"

filter_deathdata$Death_weeks <- round( filter_deathdata$Death_Days / 7.0 )
filter_deathdata$Death_months <- round( filter_deathdata$Death_Days / 30.0 )
 
print(str(filter_deathdata))  

filter_riluzole <- 
  riluzole  %>% 
  unique()


# select columns "subject_id","ALSFRS_Delta","ALSFRS_R_Total"
# filter rows with values in columns "ALSFRS_Delta","ALSFRS_R_Total"
# average value of "ALSFRS_R_Total" in a certain in moment of time
# we have several records some time

filter_alsrfs <- 
     alsrfs %>% 
     select(impotantcolumns1) %>% 
     filter(!is.na(ALSFRS_R_Total)) %>%
     filter(!is.na(ALSFRS_Delta))  %>%
     group_by(subject_id, ALSFRS_Delta) %>%
     summarise(avg_ALSFRS_R_Total = mean(ALSFRS_R_Total)) %>%
     select(impotantcolumns2) %>%
     ungroup() 
    

colnames(filter_alsrfs)[colnames(filter_alsrfs) == "avg_ALSFRS_R_Total"] = "ALSFRS_R_Total"
filter_alsrfs$ALSFRS_R_Total <- round( filter_alsrfs$ALSFRS_R_Total )

print(str(filter_alsrfs))

```
## Original statistics preparation
```{r}
# Calculate simple statistics: number of measurement (visits), day of first visit, last visit
# time between first and last visit days, weeks, months
stat_filter_alsrfs <- 
   filter_alsrfs %>%
   group_by(subject_id) %>%
   summarise( Number_visits = n(), 
              first_visit = min(ALSFRS_Delta), 
              last_visit = max(ALSFRS_Delta), 
              time_interval_days = 1+max(ALSFRS_Delta) - min(ALSFRS_Delta) ) 

stat_filter_alsrfs$time_interval_weeks <- round( stat_filter_alsrfs$time_interval_days / 7.0 )
stat_filter_alsrfs$time_interval_months <- round( stat_filter_alsrfs$time_interval_days / 30.0 )

# Add value of ALSFRS_R in the first measurement
stat_filter_alsrfs <- 
     stat_filter_alsrfs  %>%
     inner_join(filter_alsrfs, by=c("subject_id" = "subject_id", "first_visit" =  "ALSFRS_Delta")) %>%
     unique()
   
colnames(stat_filter_alsrfs)[colnames(stat_filter_alsrfs) == "ALSFRS_R_Total"] = "intial_ALSFRS_R_Total"

# Add value of ALSFRS_R in the last measurement
stat_filter_alsrfs <- 
     stat_filter_alsrfs  %>%
     inner_join(filter_alsrfs, by=c("subject_id" = "subject_id", "last_visit" =  "ALSFRS_Delta")) %>%
     unique()

colnames(stat_filter_alsrfs)[colnames(stat_filter_alsrfs) == "ALSFRS_R_Total"] = "last_ALSFRS_R_Total"

print(str(stat_filter_alsrfs))

stat_filter_alsrfs <-
  stat_filter_alsrfs  %>%
  left_join(filter_deathdata, by=c("subject_id" = "subject_id"))

print(str(stat_filter_alsrfs))

```

## Degradation, Slope
```{r}
# Changing of ALSFRS_R during all time
stat_filter_alsrfs$degradation <- stat_filter_alsrfs$intial_ALSFRS_R_Total - stat_filter_alsrfs$last_ALSFRS_R_Total

#More than one measurement + more than 1 month
stat_filter_alsrfs_op <- 
  stat_filter_alsrfs %>%
  filter(Number_visits>1) %>%
  filter(time_interval_months>0)
  

# Slope  of changing of ALSFRS_R during all time
stat_filter_alsrfs_op$slopedegradation <- stat_filter_alsrfs_op$degradation / stat_filter_alsrfs_op$time_interval_months


filter_alsrfs <- 
  filter_alsrfs %>%
  mutate(ALSFRS_Delta_weeks = ALSFRS_Delta / 7.0) 
filter_alsrfs$ALSFRS_Delta_weeks <- round( filter_alsrfs$ALSFRS_Delta_weeks )

filter_alsrfs_st <-
  filter_alsrfs %>%
  inner_join(stat_filter_alsrfs, by=c("subject_id" = "subject_id")) 


filter_alsrfs_stop <-
  filter_alsrfs %>%
  inner_join(stat_filter_alsrfs_op, by=c("subject_id" = "subject_id")) 

print(str(stat_filter_alsrfs))
print(str(stat_filter_alsrfs_op))
print(str(filter_alsrfs_st))
print(str(filter_alsrfs_stop))
```


## Correlation 

```{r}

# I have used days for correlation but maybe months is better
result_cor1 <- cor(stat_filter_alsrfs$Number_visits, stat_filter_alsrfs$time_interval_days, method = c("pearson"))
result_cor.test1 <- cor.test(stat_filter_alsrfs$Number_visits, stat_filter_alsrfs$time_interval_days, method=c("pearson"))

print(result_cor1)
print(result_cor.test1)

result_cor2 <- cor(stat_filter_alsrfs$intial_ALSFRS_R_Total, stat_filter_alsrfs$time_interval_days, method = c("pearson"))
result_cor.test2 <- cor.test(stat_filter_alsrfs$intial_ALSFRS_R_Total, stat_filter_alsrfs$time_interval_days, method=c("pearson"))

print(result_cor2)
print(result_cor.test2)

result_cor3 <- cor(stat_filter_alsrfs_op$Number_visits, stat_filter_alsrfs_op$time_interval_days, method = c("pearson"))
result_cor.test3 <- cor.test(stat_filter_alsrfs_op$Number_visits, stat_filter_alsrfs_op$time_interval_days, method=c("pearson"))

print(result_cor3)
print(result_cor.test3)

result_cor4 <- cor(stat_filter_alsrfs_op$intial_ALSFRS_R_Total, stat_filter_alsrfs_op$time_interval_days, method = c("pearson"))
result_cor.test4 <- cor.test(stat_filter_alsrfs_op$intial_ALSFRS_R_Total, stat_filter_alsrfs_op$time_interval_days, method=c("pearson"))

print(result_cor4)
print(result_cor.test4)


```

## Histogram: Number of measurement 

You can also embed plots, for example:

```{r pressure, echo=FALSE}

ggplot(stat_filter_alsrfs, aes(x = Number_visits)) + 
  geom_histogram(color="blue", fill="lightblue", binwidth=1) +
  geom_vline(aes(xintercept=mean(Number_visits)),  color="black", linetype="dashed", linewidth=1)+
  geom_vline(aes(xintercept=median(Number_visits)),  color="red", linetype="dashed", linewidth=1)


ggplot(stat_filter_alsrfs, aes(x = Number_visits)) + 
  geom_histogram(aes(y=after_stat(density)), color="blue", fill="lightblue", binwidth=1)+
  geom_density(alpha=.2, fill="#FF6666") 

```

## Histogram: Initial ALSFRS_R
```{r pressure, echo=FALSE}

ggplot(stat_filter_alsrfs, aes(x = intial_ALSFRS_R_Total)) + 
  geom_histogram(color="blue", fill="lightblue", binwidth=1) +
    geom_vline(aes(xintercept=mean(intial_ALSFRS_R_Total)),  color="black", linetype="dashed", linewidth=1) +
  geom_vline(aes(xintercept=median(intial_ALSFRS_R_Total)),  color="red", linetype="dashed", linewidth=1)


ggplot(stat_filter_alsrfs, aes(x =intial_ALSFRS_R_Total)) + 
  geom_histogram(aes(y=after_stat(density)), color="blue", fill="lightblue", binwidth=1)+
  geom_density(alpha=.2, fill="#FF6666") 

```
## Histogram: last ALSFRS_R
```{r pressure, echo=FALSE}

ggplot(stat_filter_alsrfs, aes(x = last_ALSFRS_R_Total)) + 
  geom_histogram(color="blue", fill="lightblue", binwidth=1) +
    geom_vline(aes(xintercept=mean(last_ALSFRS_R_Total)),  color="black", linetype="dashed", linewidth=1) +
  geom_vline(aes(xintercept=median(last_ALSFRS_R_Total)),  color="red", linetype="dashed", linewidth=1)


ggplot(stat_filter_alsrfs, aes(x =last_ALSFRS_R_Total)) + 
  geom_histogram(aes(y=after_stat(density)), color="blue", fill="lightblue", binwidth=1)+
  geom_density(alpha=.2, fill="#FF6666") 

```
## Histogram: degradation ALSFRS_R
```{r pressure, echo=FALSE}

ggplot(stat_filter_alsrfs, aes(x = degradation)) + 
  geom_histogram(color="blue", fill="lightblue", binwidth=1) +
    geom_vline(aes(xintercept=mean(degradation)),  color="black", linetype="dashed", linewidth=1) +
  geom_vline(aes(xintercept=median(degradation)),  color="red", linetype="dashed", linewidth=1)


ggplot(stat_filter_alsrfs, aes(x =degradation)) + 
  geom_histogram(aes(y=after_stat(density)), color="blue", fill="lightblue", binwidth=1)+
  geom_density(alpha=.2, fill="#FF6666") 

```

## Histogram: degradation ALSFRS_R (for more that one measurement) 
```{r pressure, echo=FALSE}

ggplot(stat_filter_alsrfs_op, aes(x = degradation)) + 
  geom_histogram(color="blue", fill="lightblue", binwidth=1) +
    geom_vline(aes(xintercept=mean(degradation)),  color="black", linetype="dashed", linewidth=1) +
  geom_vline(aes(xintercept=median(degradation)),  color="red", linetype="dashed", linewidth=1)


ggplot(stat_filter_alsrfs_op, aes(x =degradation)) + 
  geom_histogram(aes(y=after_stat(density)), color="blue", fill="lightblue", binwidth=1)+
  geom_density(alpha=.2, fill="#FF6666") 

```

## Histogram: slope degradation ALSFRS_R (for more that one measurement) 
```{r pressure, echo=FALSE}

ggplot(stat_filter_alsrfs_op, aes(x = slopedegradation)) + 
  geom_histogram(color="blue", fill="lightblue", binwidth=0.25) +
    geom_vline(aes(xintercept=mean(slopedegradation)),  color="black", linetype="dashed", linewidth=1) +
  geom_vline(aes(xintercept=median(slopedegradation)),  color="red", linetype="dashed", linewidth=1)


ggplot(stat_filter_alsrfs_op, aes(x = slopedegradation)) + 
  geom_histogram(aes(y=after_stat(density)), color="blue", fill="lightblue", binwidth=0.25)+
  geom_density(alpha=.2, fill="#FF6666") 

```
## Number of measurements depends on time interval (months)
```{r pressure, echo=FALSE}

ggplot(stat_filter_alsrfs, aes(x = Number_visits, y = time_interval_months)) +
  geom_point(size=2,aes(color = Number_visits))

```
## Initial ALFRS_R depends on time interval (months)
```{r pressure, echo=FALSE}

ggplot(stat_filter_alsrfs, aes(x = intial_ALSFRS_R_Total, y = time_interval_months)) +
  geom_point(size=2,aes(color = intial_ALSFRS_R_Total))

```
## Last ALFRS_R depends on time interval (months)
```{r pressure, echo=FALSE}

ggplot(stat_filter_alsrfs, aes(x = last_ALSFRS_R_Total, y = time_interval_months)) +
  geom_point(size=2,aes(color = last_ALSFRS_R_Total))

```
## degradatio ALFRS_R depends on time interval (months)
```{r pressure, echo=FALSE}

ggplot(stat_filter_alsrfs_op, aes(x = degradation, y = time_interval_months)) +
  geom_point(size=2,aes(color = degradation))

```
## scope degradation ALFRS_R depends on time interval (months)
```{r pressure, echo=FALSE}

ggplot(stat_filter_alsrfs_op, aes(x = slopedegradation, y = time_interval_months)) +
  geom_point(size=2,aes(color = slopedegradation))

```
## initial & last  ALFRS_R depends on time interval (months)
```{r pressure, echo=FALSE}

ggplot(stat_filter_alsrfs_op, aes(x = intial_ALSFRS_R_Total, y = last_ALSFRS_R_Total)) +
  geom_point(size=2,aes(color = intial_ALSFRS_R_Total))

```
## ALSFRS_R =48
```{r pressure, echo=FALSE}
p1 <- 
  filter_alsrfs_stop %>%
  filter(intial_ALSFRS_R_Total==48)

ggplot(p1, aes(x = ALSFRS_Delta_weeks, y = ALSFRS_R_Total)) +
   geom_line(aes(color=subject_id))+
   geom_point(aes(color=subject_id))
```

## ALSFRS_R =47
```{r pressure, echo=FALSE}
p1 <- 
  filter_alsrfs_stop %>%
  filter(intial_ALSFRS_R_Total==47)

ggplot(p1, aes(x = ALSFRS_Delta_weeks, y = ALSFRS_R_Total)) +
   geom_line(aes(color=subject_id))+
   geom_point(aes(color=subject_id))
```


## ALSFRS_R =46
```{r pressure, echo=FALSE}
p1 <- 
  filter_alsrfs_stop %>%
  filter(intial_ALSFRS_R_Total==46)

ggplot(p1, aes(x = ALSFRS_Delta_weeks, y = ALSFRS_R_Total)) +
   geom_line(aes(color=subject_id))+
   geom_point(aes(color=subject_id))
```